\doctitle

### 2023-05-06 conclusion.tex
Incorporate Alanna typo fixing

In this work I gave two examples of how partial knowledge can be utilized to improve performance of machine learning models. First, in Section~\ref{sec:MSR3} we showed how complex priors can be embedded into a linear mixed-effects model via a suitable proximal operator, and how the resulting algorithm can be implemented using a proximal gradient descent. We note that a complex combination of priors often yields ill-conditioned problems, and we propose a MSR3 relaxation to address it. We also prove that a solution of such relaxation approaches the solution of the original problem as the relaxation tightens. Next, we showed how partial knowledge of physics can be embedded into deep-learning-based reduced-order models. Namely, we show how to project known physical equations that govern behavior of the system in an observable space into a chosen latent space. In that space one can require that the gradient field should match the projected true dynamics. We show that this methodology leads to identification of more stable latent manifolds, and using such manifolds for forecasting leads to more accurate predictions and lesser sensitivity to noise. We also showed how the reduced-order models described above can be used for compressive sensing. We demonstrated that they can accurately reconstruct the dynamics from partial observations, significantly surpassing current state-of-the-art compressive-sensing methods. We hope that these developments pave the way to higher-quality machine learning models which effectively marry data and the humanity's hard-won expert knowledge.

\paragraph{Perspective Research Directions} Due to the limited time and resources we left many promising directions for future exploration. One such direction is extending sparse relaxed regularized regression (SR3) to generalized linear and generalized mixed-effects models. The fundamental difficulty that arises in such extension is the absence of closed-form marginal log-likelihoods. It implies that one would have three nested levels of optimization when they apply an SR3 relaxation on top of such likelihood: evaluating the likelihood, evaluating a value function, and optimizing over the value function. Re-using the recipe that we developed in Section~\ref{sec:MSR3} could help to surmount such obstacle. For example, Algorithm~\ref{alg:pgd_for_value} had two nested optimization methods: for evaluating the value function and for optimizing over it. However we showed that one can ``blend'' both levels of optimization and achieve superior performance results. The same methodology of ``blending'' could possibly be applied to more than two nested optimization loops to yield algorithms that, in certain scenarios, work faster than non-nested optimization methods.

	Our development of physics-informed reduced-order models also invites future work in several promising directions. First, a more systematic way of choosing collocation points would greatly benefit the method. In Section~\ref{sec:collocations_conditions} we define collocation points and provide criteria for a good choice of a family of colocation points. However, we ultimately leave the reader without a constructive algorithm for identifying candidate families of collocation points suitable for their problem. At the same time, certain differential equations admit families of basis functions that can span their solutions. Such basis functions frequently poses easily-computable derivatives, which makes them perfect candidates for collocation points for PINODE. Bridging the gap between this new methodology and those classic results would undoubtedly yield a fruitful line of works. Finally, one could utilize PINODE models not only in compressive sensing applications but in any application where one needs to quickly search over a large space of possible simulations. These examples include medical imaging, model discovery, online control, fast simulations, and many others.

>>>>

In this work, I gave two examples of how partial knowledge can be utilized to improve performance of machine learning models. First, in Section~\ref{sec:MSR3}, we showed how complex priors can be embedded into a linear mixed-effects model via a suitable proximal operator, and how the resulting algorithm can be implemented using a proximal gradient descent. We note that a complex combination of priors often yields ill-conditioned problems, and we propose MSR3 relaxation to address it. We also prove that a solution of such relaxation approaches the solution of the original problem as the relaxation tightens. Next, we showed how partial knowledge of physics can be embedded into deep-learning-based reduced-order models. Namely, we show how to project known physical equations that govern the behavior of the system in an observable space into a chosen latent space. In that space one can require that the gradient field match the projected true dynamics. We show that this methodology leads to identification of more stable latent manifolds, and using such manifolds for forecasting leads to more accurate predictions and less sensitivity to noise. We also showed how the reduced-order models described above can be used for compressive sensing. We demonstrated that they can accurately reconstruct the dynamics from partial observations, significantly surpassing state-of-the-art compressive-sensing methods. We hope that these developments pave the way for the creation of higher-quality machine learning models which effectively marry data and the humanity's hard-won expert knowledge.

\paragraph{Perspective Research Directions} We would like to note many promising directions for future exploration. One such direction is extending sparse relaxed regularized regression (SR3) to generalized linear and generalized mixed-effects models. The fundamental difficulty that arises in such extension is the absence of closed-form marginal log-likelihoods. It implies that one would have three nested levels of optimization when they apply an SR3 relaxation on top of such likelihood: evaluating the likelihood, evaluating a value function, and optimizing over the value function. Re-using the recipe that we developed in Section~\ref{sec:MSR3} could help to surmount such obstacle. For example, Algorithm~\ref{alg:pgd_for_value} had two nested optimization methods: for evaluating the value function and for optimizing over it. However, we showed that one can ``blend'' both levels of optimization and achieve superior performance results. The same methodology of ``blending'' could possibly be applied to more than two nested optimization loops to yield algorithms that, in certain scenarios, work faster than non-nested optimization methods.

	Our development of physics-informed reduced-order models also invites future work in several promising directions. First, a more systematic way of choosing collocation points would greatly benefit the method. In Section~\ref{sec:collocations_conditions}, we define collocation points and provide criteria for a good choice of a family of colocation points. However, we ultimately leave the reader without a constructive algorithm for identifying candidate families of collocation points suitable for their problem. At the same time, certain differential equations admit families of basis functions that can span their solutions. Such basis functions frequently have easily-computable derivatives, which makes them perfect candidates for collocation points for PINODE. Bridging the gap between this new methodology and those classic results would undoubtedly yield a fruitful line of work. Finally, one could utilize PINODE models not only in compressive sensing application, but in any application where one needs to quickly search over a large space of possible simulations. These examples include medical imaging, model discovery, online control, fast simulations, and many others.

### 2023-05-06 msr3.tex
Incorporate Alanna typo fixing

Variable selection is a fundamental problem in all regression settings. In linear regression, the LASSO method~\citep{tibshirani1996regression} and related extensions have been widely used. % for this purpose. 
However, variable selection for LMEs is complicated by the nonlinear structure and relative sparsity of the within-group data. 
While standard methods and software are available for linear regression (see e.g. \texttt{glmnet}~\cite{glmnet}), there are few open source libraries for variable selection 
for LMEs. 
Many covariates selection algorithms for LMEs have been proposed over the last 20 years (see the survey \cite{Buscemi2019Survey}), but comparison of 
these strategies and practical application remains difficult. %practical application of these strategies is difficult. 
%Most solutions share the goal of improving selection quality but vary significantly in implementation details. 
Approaches vary by choice of likelihood (e.g. marginal, restricted, or h- likelihood),  
regularizer~(e.g. $\ell_1$~\citep{Krishna2008} or SCAD~\cite{ibrahim2011fixed}), and information criteria \citep{Vaida2005,Ibrahim2011}. 
%and 
%sometimes  estimate effects in stages \citep{Krishna2008}. 
Implementations vary as well, typically using regularizer-specific local quadratic approximations to apply
solution methods for smooth problems (Newton-Raphson, EM, sequential least squares) to fit the original nonsmooth model. 
All of these decisions make it  difficult to compare and evaluate performance of available 
variable selection strategies and to determine which method is best suited for a given task. 
This challenge  is exacerbated by the absence of standardized datasets 
%for different experimental designs
and open source libraries for each method. 
Our main practical goal to fill this gap by developing a unified methodological framework that
accommodates a wide variety of variable selection strategies based on a set
of easily implementable regularizers, and 
made available in an open source library, \texttt{pysr3}\footnote{https://github.com/aksholokhov/pysr3} that is 
easy to use and to compare different methods. All experiments in the paper can be reproduced 
using \texttt{pysr3} and code in the reproducibility guide\footnote{https://github.com/aksholokhov/msr3-paper}.

@@

%Roadmap
We begin in Section \ref{sec:LMEM} by giving a precise description of the LME
model and set the notation for the remainder of the paper. 
This is followed by a brief discussion of prior work on LMEs.
In Section \ref{sec:pgd_methods} we present our algorithms for the LME model 
starting with the naive PGD algorithm. This is followed by a description of the 
variable splitting technique used in \cite{Zheng2019SR3} to incorporate global 
variation information on the likelihood function into the direction finding subproblem
for the PGD algorithm. 
%This lays the foundation for the MSR3 algorithm. 
Next we tackle the problem of how to approximate the 
resulting optimal value function $u_\eta$ and its gradient where $\eta>0$ is the 
decoupling parameter. As noted, this is done using variable metric techniques and 
and interior point technology. We conclude Section \ref{sec:pgd_methods} 
with a discussion of the
MSR3 and MSR3-fast algorithms. In Section \ref{sec:applications} we discuss how
the underlying algorithmic parameters are set and test the algorithm on both 
simulated problems and a problem with real data. In Section~\ref{sec:theory-paper} we develop theoretical underpinnings of the proposed extension, including consistency results, variational properties, implementability of optimization methods, and convergence results for both MSR3 and MSR3-fast. Section~\ref{sec:software-paper} outlines the software implementation of MSR3 and MSR3-fast in \texttt{pysr3} package. The chapter is concluded in Section \ref{sec:discussion} with a brief discussion of the contributions.

@@

This paper focuses the case where $\Gamma$ is diagonal, 
  (often referred to as \textit{the diagonal setup}) 
 and all $\Lambda_i$ are known (see \eqref{eq:lme_diagonal_setup}), 
following the meta-analysis use-case~\citep{zheng2021trimmed} that is widely 
employed in epidemiological studies~\cite{murray2020global}. 
While the proposed approach can be extended to the non-diagonal case, we leave it for future work, save for a brief discussion in Section~\ref{sec:applications}.

@@

    In this paper, we developed and implemented a first-order variable selection framework for LMEs that handles convex and nonconvex regularizers. We also showed that 
    the MSR3 relaxation \eqref{eq:value_function_optimization} 
%a regularizer-agnostic transformation of the likelihood, 
improves the covariates selection accuracy of a wide group of popular sparsity-promoting regularizers. The fact that the relaxation improves accuracy, rather than just serving as a means to {numertical efficiency, %solve the original problem,
     is very interesting and deserves future study.} % result that can be studied in future work.

>>>>

Variable selection is a fundamental problem in all regression settings. In linear regression, the LASSO method~\citep{tibshirani1996regression} and related extensions have been widely used. % for this purpose. 
However, variable selection for LMEs is complicated by the nonlinear structure and relative sparsity of the within-group data. 
While standard methods and software are available for linear regression (see e.g. \texttt{glmnet}~\cite{glmnet}), there are few open source libraries for variable selection 
for LMEs. 
Many covariates selection algorithms for LMEs have been proposed over the last 20 years (see the survey \cite{Buscemi2019Survey}), but comparison of 
these strategies and practical application remains difficult. %practical application of these strategies is difficult. 
%Most solutions share the goal of improving selection quality but vary significantly in implementation details. 
Approaches vary by choice of likelihood (e.g. marginal, restricted, or h- likelihood),  
regularizer~(e.g. $\ell_1$~\citep{Krishna2008} or SCAD~\cite{ibrahim2011fixed}), and information criteria \citep{Vaida2005,Ibrahim2011}. 
%and 
%sometimes  estimate effects in stages \citep{Krishna2008}. 
Implementations vary as well, typically using regularizer-specific local quadratic approximations to apply
solution methods for smooth problems (Newton-Raphson, EM, sequential least squares) to fit the original nonsmooth model. 
All of these decisions make it  difficult to compare and evaluate performance of available 
variable selection strategies and to determine which method is best suited for a given task. 
This challenge  is exacerbated by the absence of standardized datasets 
%for different experimental designs
and open source libraries for each method. 
Our main practical goal to fill this gap by developing a unified methodological framework that
accommodates a wide variety of variable selection strategies based on a set
of easily implementable regularizers, and 
made available in an open source library, \texttt{pysr3}\footnote{https://github.com/aksholokhov/pysr3} that is 
easy to use and to compare different methods. All experiments in this work can be reproduced 
using \texttt{pysr3} and code in the reproducibility guide\footnote{https://github.com/aksholokhov/msr3-paper}.

@@

%Roadmap
We begin in Section \ref{sec:LMEM} by giving a precise description of the LME
model and set the notation for the remainder of the section. 
This is followed by a brief discussion of prior work on LMEs.
In Section \ref{sec:pgd_methods} we present our algorithms for the LME model 
starting with the naive PGD algorithm. This is followed by a description of the 
variable splitting technique used in \cite{Zheng2019SR3} to incorporate global 
variation information on the likelihood function into the direction finding subproblem
for the PGD algorithm. 
%This lays the foundation for the MSR3 algorithm. 
Next we tackle the problem of how to approximate the 
resulting optimal value function $u_\eta$ and its gradient where $\eta>0$ is the 
decoupling parameter. As noted, this is done using variable metric techniques and 
and interior point technology. We conclude Section \ref{sec:pgd_methods} 
with a discussion of the
MSR3 and MSR3-fast algorithms. In Section \ref{sec:applications} we discuss how
the underlying algorithmic parameters are set and test the algorithm on both 
simulated problems and a problem with real data. In Section~\ref{sec:theory-paper} we develop theoretical underpinnings of the proposed extension, including consistency results, variational properties, implementability of optimization methods, and convergence results for both MSR3 and MSR3-fast. Section~\ref{sec:software-paper} outlines the software implementation of MSR3 and MSR3-fast in \texttt{pysr3} package. The chapter is concluded in Section \ref{sec:discussion} with a brief discussion of the contributions.

@@

Our work focuses the case where $\Gamma$ is diagonal, 
  (often referred to as \textit{the diagonal setup}) 
 and all $\Lambda_i$ are known (see \eqref{eq:lme_diagonal_setup}), 
following the meta-analysis use-case~\citep{zheng2021trimmed} that is widely 
employed in epidemiological studies~\cite{murray2020global}. 
While the proposed approach can be extended to the non-diagonal case, we leave it for future work, save for a brief discussion in Section~\ref{sec:applications}.

@@

    In this work, we developed and implemented a first-order variable selection framework for LMEs that handles convex and nonconvex regularizers. We also showed that 
    the MSR3 relaxation \eqref{eq:value_function_optimization} 
%a regularizer-agnostic transformation of the likelihood, 
improves the covariates selection accuracy of a wide group of popular sparsity-promoting regularizers. The fact that the relaxation improves accuracy, rather than just serving as a means to {numertical efficiency, %solve the original problem,
     is very interesting and deserves future study.} % result that can be studied in future work.

### 2023-05-06 phd_thesis.tex
Address multiple TODO items in MSR3 chapter
Incorporate Alanna typo fixing

My abstract goes here. I will probably write it at the very end.

>>>>

This work summarizes two projects focused on incorporating prior knowledge into machine learning models. In the first project, we developed universal feature selection methods for linear mixed-effect models. Namely, we extended Sparse Relaxed Regularized Regression (SR3) to Linear Mixed-Effect (LME) likelihood and showed how one can minimize such likelihood with proximal gradient descent. We also developed theoretical underpinnings of the proposed extension, including consistency results, variational properties, implementability of optimization methods, and convergence results. In particular, we provided convergence analyses for a basic implementation of SR3 for LME and an accelerated hybrid algorithm. Numerical results show the utility and speed of these algorithms on realistic simulated datasets. Finally, we provide an open-source implementation of both algorithms in an open-source python package \texttt{pysr3}. This package offers complete compatibility with \texttt{scikit-learn}, so all \texttt{pysr3} models can be used in pipeline with classic modelling blocks such as data pre-processors, randomized grid search, cross-validation, and quality metrics.

The second line of work develops a framework for training Reduced-Order Models (ROMs) with Physics-Informed Neural Ordinary Differential Equations (PINODE). In particular, we illustrate how one can utilize a classic technique of collocation points to transfer knowledge from a known equation to a model that approximates solutions of that equation. We show that the addition of our physics-informed loss allows for exceptional data supply strategies that improve the performance of ROMs in data-scarce settings, where training high-quality data-driven models is impossible. The resulting ROMs extrapolate forward in time considerably more effectively,  perform better for unseen initial conditions, and exhibit less sensitivity to noise. Finally, we show how such ROMs can be used as strong regularizers in single-pixel imaging (SPI) enabling the reduction of sampling-per-frame rate by an order of magnitude relative to the state-of-the-art algorithms.

### 2023-05-06 pinode.tex
Incorporate Alanna typo fixing

\section{Introduction}
\edit{Write a big-picture intro to machine learning for physics}
\edit{Replace mentions of paper with work}
\subsection{Related Work}
\edit{Split original intro into prior works paragraphs: DL for Ph, ROMs, PIML}
\edit{Add physics-informed ML overview (PINN, DeepONet, NFO)}
        Forecasting the behavior of a large-scale real-world system directly from first principles often requires solving highly-nonlinear governing equations such as high-dimensional ordinary differential equations (ODEs) or partial differential equations (PDEs). 
        High-fidelity simulations of such dynamical systems can become intractable, especially if an online control algorithm requires multiple forecasts per second using a low-powered embedded device \cite{rowley2017model,lucia2004reduced,benner2015survey}. A situation like this arises, for example, when a smart heating, ventilation, and air conditioning (HVAC) system attempts to optimize the temperature distribution of the air in a room using only partial measurements \cite{farahmand2016learning,nabi2022robust}. At the time of writing this paper, such systems are incapable of real-time complex simulations, but they can already run low-dimensional pre-trained models, which invites the development of high-quality reduced order models (ROMs) \cite{otterness2017evaluation}. Therefore, ROMs are essential for enabling the design optimization, uncertainty propagation, predictive modeling, and control for such dynamical systems \cite{brunton2022data,kutz2016dynamic,rowley2017model,jones2020characterising}

@@

        In this paper, we employ autoencoders to perform nonlinear model reduction along with NODE in the latent space to model complex and nonlinear dynamics. We choose Neural ODEs in the latent space dynamics representation because of their ability to model highly non-linear dynamics, which is especially important when applications limit the size of the latent space dimension. Our goal is to reduce the demand for training data and improve the overall forecasting stability under challenging training conditions. To that end, we build on ideas from classical collocation methods of numerical analysis to embed knowledge from a known governing equation into the latent-space dynamics of a ROM, as described in Section 2. In Section 3, we show that the addition of our physics-informed loss allows for exceptional data supply strategies that improves the performance of ROMs in data-scarce settings, where training high-quality data-driven models is impossible. We demonstrate that such an approach not only reduces the need for large training data-sets and produces highly-accurate and long-term stable models, but also leads to the discovery of more compact latent spaces, which is especially important for applications in compressed sensing and control.

@@

\begin{align}
    \label{eq:loss_data_driven}
    \LL^{data}_\theta & = \frac{1}{2\sigma^2}\sum_{i = 1}^k \left[\frac{\omega_1}{p}\sum_{j=1}^p\left\|\bd{x_i}(t_j) - \psi_\theta(\phi_\theta(\bd{x_i}(t_j)))\right\|^2\right. + \\
     & + \left.\frac{\omega_2}{p}\sum_{j=1}^p \left\|\psi_\theta\left(\phi_\theta(\bd{x_i}(t_1)) + \int_{t_1}^{t_j}h(z(t))dt\right) - \bd{x_i}(t_j)\right\|^2 \right]
\end{align}
where $\sigma$ is the standard deviation of the observation noise. We note that each trajectory $\bd{x}_i$ may be captured over its own time-frame and may use a distinct, possibly non-uniform, step-size, in which case the loss function should be modified accordingly\footnote{The implementation is affected only in evaluating the integral in~(\ref{eq:loss_data_driven}). This part is handled by \texttt{torchdiffeq}~\cite{chen2018neural} library, which supports non-uniform time-frames within a batch}. To simplify the notation, without loss of generality, in the rest of the paper we assume that all trajectories are recorded over the same time-frame with the same uniform step-size. To forecast the behavior of the system in the latent space, we apply the technique of Neural Ordinary Differential Equations (Neural ODEs or NODEs)\cite{chen2018neuralode}, which utilizes the adjoint sensitivity method to back-propagate the gradients through the integral in~(\ref{eq:loss_data_driven}). Neural ODEs have demonstrated a better ability to model highly non-linear dynamics compared to linear models when the dimensionality of the dynamics variable is limited. This is especially useful in applications where the size of the latent space dimension needs to be small~\cite{lee2020model,gin2021deep,champion2019data,kim2019deep}.

@@

The coefficients $\omega_i$ are hyper-parameters which need to be tuned using a validation dataset. However, in all experiments of this paper we set $\omega_i$ to be either $0$ or $1$, and we balance $\LL^{physics}_\theta$ and $\LL^{data}_\theta$ the choice of samples in a batch of training data. Specifically, we set the number of collocation points per batch $N_{batch}$ to be equal to the number of trajectories per batch $k_{batch}$ times the number of time-steps$T$: $N_{batch} = Tk_{batch}$. In this way both $\LL^{physics}_\theta$ and $\LL^{data}_\theta$ represent the loss for $Tk_{batch}$ snapshots of the system, providing on average a similar contribution of information to the overall loss function. More laborious approaches of hyper-parameter tuning did not yield sufficient systematic advantage to justify the labour compared to this simple strategy.

@@

Figure~(\ref{fig:burger_noise}) shows that in the high noise setting, the error of purely data-driven models (DMD and PINODE Data-Driven) grows unbounded, whereas the performance of the hybrid model converges to the performance of the Physics-Informed model as the noise level increases. We hypothesise that such behavior is due to the second part~($\LL_{\theta}^{data}$) of the combined loss~(Eq.~\ref{eq:loss_combined}) turns into noise, and so its derivative also turns into noise.
\begin{equation}
    \nabla \LL_{\theta} = \underbrace{\nabla\LL_{\theta}^{physics}}_{\text{informative}} + \underbrace{\nabla\LL_{\theta}^{data}}_{\text{noise}}
\end{equation}
Thus, one can think about optimizing a hybrid model~(\ref{eq:loss_combined}) as about training a Physics-Informed model~(\ref{eq:loss_physics_informed}) using a noisy gradient descent with a fixed-variance noise. From the optimization literature \cite{friedlander2012hybrid,patel2021global,shapiro2021lectures} we know that, under certain conditions, such SGD converges to a neighbourhood of a local minimum of its loss (in this case $\LL_{\theta}^{physics}$) with high probability. So instead of diverging, a hybrid model turns into a Physics-Informed model; where the latter works as a performance safeguard in the high-noise regime.  On the right hand-side of Figure~(\ref{fig:burger_noise}), we show an example of the prediction performance of each of the models described above. The data-driven and hybrid models yield visually similar solutions when $\sigma = 10^-3$. However, the former provides inadequate performance when the data is dominated by noise, whereas a hybrid model in this regime produces a solution that is visually similar to the one that the Physics-Informed model produces. A more rigorous analysis of this phenomenon seems possible but lies outside of the scope of this paper.

\section{Discussion and Conclusions}

In this work, we demonstrated how a collocation point-based technique can improve the performance of an emerging class of continuous-time physics-informed neural-network based reduced-order models. First, we demonstrated that the incorporation of collocation points in training data can ``cover the gaps'' in training trajectories and inform the model about underrepresented basins of attraction. Such an approach alleviates the demand for large volumes of data that is common in network-based models, which is crucial in applications where data is scarce and expensive. Second, the physics-informed loss may work as a safeguard, providing a noise-free source of underlying dynamics.
Third, collocation points can stabilize the model's long-term predictions, allowing for accurate forecasting far beyond the training time horizon. Finally, together with using a NODE-based non-linear latent dynamics, adding physics-informed loss leads to the discovery of more compact latent space representations that also yield more accurate models. Simultaneous stability and compactness is especially important if one aims to use models together with compressive sensing and control algorithms. With respect to the computational complexity, we note that adding $Tk$ collocation points to the training imposes less of a computational burden than adding $k$ data trajectories because collocation points do not require computing integrals forward in time as in the case of data trajectories.

One clear limitation of the current work is that the choice of an efficient collocation family is a design decision that a practitioner makes. The authors believe that such decisions can be automated by adopting existing approaches from classic works on numerical approximations of PDEs, which we leave for future research. Another automation that prompts future research is deriving efficient ways of sampling collocation points, possibly via applying modern adaptive learning techniques \cite{subramanian2022adaptive}. Finally, although Section~\ref{sec:burger_noise} provides some rationale for why one may expect robustness of Hybrid models under noise, the authors believe that a more rigorous analysis is possible; particularly one that provides conditions under which such robustness is guaranteed.

>>>>

\section{Introduction}
\edit{Write a big-picture intro to machine learning for physics}
\subsection{Related Work}
\edit{Split original intro into prior works paragraphs: DL for Ph, ROMs, PIML}
\edit{Add physics-informed ML overview (PINN, DeepONet, NFO)}
        Forecasting the behavior of a large-scale real-world system directly from first principles often requires solving highly-nonlinear governing equations such as high-dimensional ordinary differential equations (ODEs) or partial differential equations (PDEs). 
        High-fidelity simulations of such dynamical systems can become intractable, especially if an online control algorithm requires multiple forecasts per second using a low-powered embedded device \cite{rowley2017model,lucia2004reduced,benner2015survey}. A situation like this arises, for example, when a smart heating, ventilation, and air conditioning (HVAC) system attempts to optimize the temperature distribution of the air in a room using only partial measurements \cite{farahmand2016learning,nabi2022robust}. At the time of writing this document, such systems are incapable of real-time complex simulations, but they can already run low-dimensional pre-trained models, which invites the development of high-quality reduced order models (ROMs) \cite{otterness2017evaluation}. Therefore, ROMs are essential for enabling the design optimization, uncertainty propagation, predictive modeling, and control for such dynamical systems \cite{brunton2022data,kutz2016dynamic,rowley2017model,jones2020characterising}

@@

        In this work, we employ autoencoders to perform nonlinear model reduction along with NODE in the latent space to model complex and nonlinear dynamics. We choose Neural ODEs in the latent space dynamics representation because of their ability to model highly non-linear dynamics, which is especially important when applications limit the size of the latent space dimension. Our goal is to reduce the demand for training data and improve the overall forecasting stability under challenging training conditions. To that end, we build on ideas from classical collocation methods of numerical analysis to embed knowledge from a known governing equation into the latent-space dynamics of a ROM, as described in Section 2. In Section 3, we show that the addition of our physics-informed loss allows for exceptional data supply strategies that improves the performance of ROMs in data-scarce settings, where training high-quality data-driven models is impossible. We demonstrate that such an approach not only reduces the need for large training data-sets and produces highly-accurate and long-term stable models, but also leads to the discovery of more compact latent spaces, which is especially important for applications in compressed sensing and control.

@@

\begin{align}
    \label{eq:loss_data_driven}
    \LL^{data}_\theta & = \frac{1}{2\sigma^2}\sum_{i = 1}^k \left[\frac{\omega_1}{p}\sum_{j=1}^p\left\|\bd{x_i}(t_j) - \psi_\theta(\phi_\theta(\bd{x_i}(t_j)))\right\|^2\right. + \\
     & + \left.\frac{\omega_2}{p}\sum_{j=1}^p \left\|\psi_\theta\left(\phi_\theta(\bd{x_i}(t_1)) + \int_{t_1}^{t_j}h(z(t))dt\right) - \bd{x_i}(t_j)\right\|^2 \right]
\end{align}
where $\sigma$ is the standard deviation of the observation noise. We note that each trajectory $\bd{x}_i$ may be captured over its own time-frame and may use a distinct, possibly non-uniform, step-size, in which case the loss function should be modified accordingly\footnote{The implementation is affected only in evaluating the integral in~(\ref{eq:loss_data_driven}). This part is handled by \texttt{torchdiffeq}~\cite{chen2018neural} library, which supports non-uniform time-frames within a batch}. To simplify the notation, without loss of generality, in the rest of the section we assume that all trajectories are recorded over the same time-frame with the same uniform step-size. To forecast the behavior of the system in the latent space, we apply the technique of Neural Ordinary Differential Equations (Neural ODEs or NODEs)\cite{chen2018neuralode}, which utilizes the adjoint sensitivity method to back-propagate the gradients through the integral in~(\ref{eq:loss_data_driven}). Neural ODEs have demonstrated a better ability to model highly non-linear dynamics compared to linear models when the dimensionality of the dynamics variable is limited. This is especially useful in applications where the size of the latent space dimension needs to be small~\cite{lee2020model,gin2021deep,champion2019data,kim2019deep}.

@@

The coefficients $\omega_i$ are hyper-parameters which need to be tuned using a validation dataset. However, in all experiments of this section we set $\omega_i$ to be either $0$ or $1$, and we balance $\LL^{physics}_\theta$ and $\LL^{data}_\theta$ the choice of samples in a batch of training data. Specifically, we set the number of collocation points per batch $N_{batch}$ to be equal to the number of trajectories per batch $k_{batch}$ times the number of time-steps$T$: $N_{batch} = Tk_{batch}$. In this way both $\LL^{physics}_\theta$ and $\LL^{data}_\theta$ represent the loss for $Tk_{batch}$ snapshots of the system, providing on average a similar contribution of information to the overall loss function. More laborious approaches of hyper-parameter tuning did not yield sufficient systematic advantage to justify the labour compared to this simple strategy.

@@

Figure~(\ref{fig:burger_noise}) shows that in the high noise setting, the error of purely data-driven models (DMD and PINODE Data-Driven) grows unbounded, whereas the performance of the hybrid model converges to the performance of the Physics-Informed model as the noise level increases. We hypothesise that such behavior is due to the second part~($\LL_{\theta}^{data}$) of the combined loss~(Eq.~\ref{eq:loss_combined}) turns into noise, and so its derivative also turns into noise.
\begin{equation}
    \nabla \LL_{\theta} = \underbrace{\nabla\LL_{\theta}^{physics}}_{\text{informative}} + \underbrace{\nabla\LL_{\theta}^{data}}_{\text{noise}}
\end{equation}
Thus, one can think about optimizing a hybrid model~(\ref{eq:loss_combined}) as about training a Physics-Informed model~(\ref{eq:loss_physics_informed}) using a noisy gradient descent with a fixed-variance noise. From the optimization literature \cite{friedlander2012hybrid,patel2021global,shapiro2021lectures} we know that, under certain conditions, such SGD converges to a neighbourhood of a local minimum of its loss (in this case $\LL_{\theta}^{physics}$) with high probability. So instead of diverging, a hybrid model turns into a Physics-Informed model; where the latter works as a performance safeguard in the high-noise regime.  On the right hand-side of Figure~(\ref{fig:burger_noise}), we show an example of the prediction performance of each of the models described above. The data-driven and hybrid models yield visually similar solutions when $\sigma = 10^-3$. However, the former provides inadequate performance when the data is dominated by noise, whereas a hybrid model in this regime produces a solution that is visually similar to the one that the Physics-Informed model produces. A more rigorous analysis of this phenomenon seems possible but lies outside of the scope of this work.
